Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy

### 2017-11-04 aggregation/agg-JIE.tex
Late afternoon read through

  \maketitle
  \emph{For submission to the Journal of Industrial Ecology; October 2017}

@@

  {\red http://pubs.acs.org/doi/pdf/10.1021/acs.est.5b01558}

@@

This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.  A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.  Once equipped with the PSM, the reviewer or data user can potentially go much further.  The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports.  Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.  Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to several background databases without adding any additional requirements.

>>>>

  \maketitle
  \emph{For submission to the Journal of Industrial Ecology; November 2017}

@@

%  {\red http://pubs.acs.org/doi/pdf/10.1021/acs.est.5b01558}

@@

This paper presents a possible solution by providing a mathematical formalization of the foreground of an LCA study and a functional specification for disclosing the PSM.  A data user or critical reviewer with access to the six disclosure elements, and also adequate access to background datasets and elementary flows used in the model, can reproduce the LCA computation in Eq.~\ref{eqn:study}, and ensure the reported result is correct.  Once equipped with the PSM, the reviewer or data user can potentially go much further.  The graph structure encoded in the foreground matrices can be automatically rendered as a process-flow diagram, supplementing or supplanting the hand-crafted system diagrams that currently accompany ISO reports.  Reviewers or data users who are able to reproduce the PSM from the disclosure can also modify it, adding or removing elements, altering dataset selections, and applying other impact assessment methods.

### 2017-11-04 aggregation/agg-abstract.tex
Late afternoon read through

\begin{abstract}
  Many of the challenges facing %interpretation and
  knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of life cycle inventory models.  This article presents a framework for computing aggregation results in LCA $\ldots$
\end{abstract}

>>>>

\begin{abstract}
  Many of the challenges facing knowledge synthesis from life cycle assessment (LCA) studies stem from the inability of study authors and readers to formally agree on the structure and content of the product system models used to perform LCA computations.
  This article presents a framework for distinguishing the portions of an LCA study that are attributable to the design of the model from

  isolating the

  computing aggregation results in LCA $\ldots$
\end{abstract}

### 2017-11-04 aggregation/agg-basic-lca.tex
Late afternoon read through

Because the objective of the framework is to validate a computation, we begin by providing a mathematical formulation of the LCA problem.  An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.  The task of LCA can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.  If the processes are assumed to be linear, the problem can be written in terms of matrix algebra.  There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Heijungs2002, Pauliuk_2015_framework}.  Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_JLCA_2007}.  That formulation may be stated:
\begin{equation}
s = \mathbf{e}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
\label{eqn:leontief}
\end{equation}
Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values) per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.  In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.  For simplicity, we will consider the computation of a single impact result.  However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.

@@

This formulation is equivalent to the implementations of all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.  However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are the same for all users.  Thus exporting $A$ and $B$ violates our authority and primacy objectives.  Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.   As a consequence, the matrix formulation in Eq.~\ref{eqn:leontief} is unsuitable as a disclosure framework.

>>>>

We begin by providing a mathematical formulation of the LCA problem.  An LCA computation is performed over a collection of linked unit processes, each of which describes a single operation or reference product.  The task of procedure can be described in two steps: making an inventory of environmental interventions; and determining their environmental impacts.  If the processes are assumed to be linear and time-invariant, the problem can be written in terms of matrix algebra.  There are two common formulations: the ``technology matrix'' approach and the ``input-output'' or Leontief approach \citep{Heijungs2002, Pauliuk_2015_framework}.  Although the two models are understood to be equivalent, the input-output approach is computationally more efficient in some applications because it lends itself to iterative solution \citep{Peters_JLCA_2007}.  That formulation may be stated:
\begin{equation}
s = \mathbf{e}^T \cdot B \cdot \left(I - A\right)^{-1} \cdot \mathbf{y}
\label{eqn:leontief}
\end{equation}
Here $A$ represents the \emph{direct requirements matrix}, where each column of values reports the necessary inputs (positive values) and generated non-reference outputs (negative values)%\footnote{Whether negative values are present or allowed depends on the particular construct or co-production method being used. \cite{Majeau_Bettez_2014}}
per unit of a process's reference output; $B$ is the environmental intervention or emission matrix; and $\mathbf{e}$ is a column vector of characterization factors for the environmental emissions with respect to a given impact assessment method.  In this model, $\mathbf{y}$ is the externally specified final demand, which is the product or service whose delivery is to be assessed, and $s$ is the numerical impact score or category indicator.  For simplicity, we will consider the computation of a single impact result.  However, it is straightforward to imagine the more typical case, in which $E$ is a matrix of characterization vectors, and $\mathbf{s}$ is a vector of results.

@@

This formulation implemented in all major LCA software systems, some of which provide the capacity to export the $A$ and $B$ matrices for a given study model.  However, the contents of commercial LCIDBs are proprietary and subject to licensing restrictions, and moreover are not usually modified by study authors.  Thus exporting $A$ and $B$ does not meet our authority and primacy objectives.  Additionally, replicating the entire LCIDB is cumbersome because of the large size of the matrices.   As a consequence, the matrix formulation in Eq.~\ref{eqn:leontief} is unsuitable as a disclosure framework.

### 2017-11-04 aggregation/agg-disclosure.tex
Late afternoon read through

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together can be considered a complete representation of the PSM.

@@

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references in \ref{itm:bg} and \ref{itm:em};
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes only information for which the study is the \textbf{primary} source.
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.  These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.  However, the the challenges of operationalizing this requirement are complex.  Some implications on the information required for a successful disclosure are considered in the Discussion.

>>>>

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$ (or their product $\mathbf{e}\times B_x$) are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together can be considered a complete representation of the PSM.

@@

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references in \ref{itm:bg} and \ref{itm:em};
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes only information for which the study is the \textbf{primary} source.  
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.  These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.  However, the the challenges of operationalizing this requirement are complex (see Discussion and Supplementary materials).  %Some implications on the information required for a successful disclosure are considered in the Discussion.

### 2017-11-04 aggregation/agg-intro.tex
Finish discussion?
afternoon work- cites; semantics; editing
Late afternoon read through

An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.  But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.  For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.  But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.  When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.  ISO 14071, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional \citep{iso14071}.

A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.  Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.  Moreover, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.  The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.

>>>>

An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.  But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.  For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.  But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.  When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.  \cite{iso14071}, which provides guidelines on critical review, emphasizes that review of the PSM or individual datasets is optional.

%A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.
Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.  Moreover, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.  The purpose of this paper is to advance a framework for the formal description of product system models that would provide robust transparency and reproducibility of computed results.

### 2017-11-04 aggregation/agg-lca-study.tex
Late afternoon read through

In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.  To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.  Because background databases are prepared independently of any study, their computational results do not vary across the studies in which they are used.  However, the foreground systems themselves cannot be computed without the background.  Put another way, the foreground depends on the background but the background does not depend on the foreground.  This condition allows the processes in an LCIDB to be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.

@@

\input{fig_foreground}
Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.  An LCA study augmented with foreground information can be written in block-triangular form:
\begin{equation}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];\;\;\;
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
\label{eqn:foreground}
\end{equation}
Here, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions. The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.

For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.  This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.  However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).

>>>>

In practice, most LCA studies contain foreground information that is not present in a background database, and LCA software systems generally allow users to augment the standard LCIDBs with foreground models.  To meet our disclosure objectives, it is important to be able to distinguish foreground content from background content in the computation.  Background databases are prepared independently of any study and do not typically vary across the studies in which they are used.  However, the foreground systems themselves cannot be computed without the background.  Put another way, the foreground depends on the background but the background does not depend on the foreground.  This condition allows the processes in an LCIDB to be partially ordered to group foreground processes together \citep{Kuczenski_JLCA_2015}.

@@

\input{fig_foreground}
Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.  An LCA study augmented with foreground information can be written in block-triangular form:
\begin{equation}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];\;\;\;
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
\label{eqn:foreground}
\end{equation}
Here, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions. The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance. However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2015}).

%For clarity, I will call thein which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be an

The augmented $\tilde{A}$ and $\tilde{B}$, makes up an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.  This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.

### 2017-11-04 aggregation/agg-objective.tex
Late afternoon read through

The task of computing life cycle impact assessment (LCIA) results from a product system model can be described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.  Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.  
In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration, from a single-operation unit process to a full ISO-style product LCA.

@@

@@

On the basis of transparency, the framework should meet the following objectives:
\begin{itemize}
\item \textbf{Computability}. The result should be computed using only the disclosure, plus resources available independently to the reader.  The computation should follow the consensus understanding of process LCA computation.
\item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
\end{itemize}
One key objective of disclosure is to inspect the work of the author.  Consequently, the disclosure should be made up of the author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.  The purpose of this is twofold: to focus the efforts of the reader on the work of the author, and to enable the reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
\begin{itemize}
\item \textbf{Authority}.  The disclosure should include only information for which the author is responsible.
\item \textbf{Primacy}.  The disclosure should exclude information that is made available by a third party.
\end{itemize}
In addition, because of the ubiquity of confidential and sensitive information in LCA, any operational disclosure framework must also meet the following:
\begin{itemize}
\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.
\end{itemize}

>>>>

The task of computing life cycle impact assessment (LCIA) results from a product system model can be described as an ``aggregation,'' in which the data in large, sparse models are combined through linear algebra operations into a small set of numeric scores.  Although several equivalent methods for computing aggregated LCI and/or LCIA results have been described in the literature \citep{Suh2005a, Peters_JLCA_2007}, the computational representation of and exchange of product system models, and the generation of aggregated results, remains software-specific.  
In the context of this paper, an ``LCA Study'' should be understood as any collection of inventory data under consideration as a group, from a single-operation unit process to a full ISO-style product LCA.

@@

On the basis of transparency, the framework should meet the following objectives:
\begin{itemize}
\item \textbf{Computability}. The result should be computed using only the disclosure, plus resources available independently to the reader.  The computation should follow the consensus understanding of process LCA computation.
\item \textbf{Completeness / Minimality}.  The computation should use all of the information provided in the disclosure.
\item \textbf{Reproducibility}.  The result obtained by the reader should match the result provided by the author.
\end{itemize}
One key objective of disclosure is to inspect the work of the author.  Consequently, the disclosure should be made up of the author's own work product and should exclude, for instance, data that were drawn from a reference database or other independent source and used in an unmodified form.  The purpose of this is twofold: to focus the efforts of the reader on the work of the author, and to enable the reader to obtain and inspect the third-party data independently, thus ensuring its integrity.
\begin{itemize}
\item \textbf{Authority}.  The contents of the disclosure should authentically represent the outcomes of the author's labor in constructing the PSM.
\item \textbf{Primacy}.  The disclosure should exclude information that is made available by a third party.  (It must, however, still include previously-published data that were re-implemented by the author in the study.)
\end{itemize}
In addition, because of the ubiquity of confidential and sensitive information in LCA, any operational disclosure framework must also meet the following:
\begin{itemize}
\item \textbf{Privacy}.  The disclosure should protect information that is held in confidence by the author.
\end{itemize}

@@

### 2017-11-04 aggregation/agg-pagg.tex
Late afternoon read through

The $A_d$ and $B_f$ matrices are then partitioned into two components that sum to the original:
\begin{equation}\begin{aligned}
  A_d =\; & A_{d,pub} + A_{d,priv} \\
  B_f =\; & B_{f,pub} + B_{f,priv}
  \label{eqn:partition}
  \end{aligned}
  \end{equation}
This is illustrated in Fig.~\ref{fig:private}b.  Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$, while the public portions remain disaggregated:
\begin{align}
  s =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
\end{align}
Any foreground nodes that are completely contained within the private partition can be omitted from the disclosed foreground.  The aggregation result can then be included as a separate foreground node with a unit weight, and the aggregated inventory vector included as the corresponding column of $B_f$.  This is illustrated in Fig.~\ref{fig:private}c.  The

>>>>

The $A_d$ and $B_f$ matrices are then partitioned into two components that sum to the original:
\begin{equation}\begin{aligned}
  A_d =\; & A_{d,pub} + A_{d,priv} \\
  B_f =\; & B_{f,pub} + B_{f,priv}
  \label{eqn:partition}
  \end{aligned}
  \end{equation}
This is illustrated in Fig.~\ref{fig:private}b.  Substituted back into Eq.~\ref{eqn:study}, the private portions are aggregated into a private life cycle inventory vector $\mathbf{b}_{priv}$, while the public portions remain disaggregated:
\begin{align}
  s =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_{f,priv}) + B_x\times (A_{d,pub} + A_{d,priv})\right)\times\tilde{\mathbf{x}}\\
  =\; & \mathbf{e}^T\times\left((B_{f,pub} + B_x\times A_{d,pub})\times\tilde{\mathbf{x}} + \mathbf{b}_{priv}\right)\label{eqn:private}\\
  \mathbf{b}_{priv} =\; & (B_{f,priv} + B_x\times A_{d,priv})\times \tilde{\mathbf{x}}
\end{align}
Any foreground nodes that are completely contained within the private partition can be omitted from the disclosed foreground.  The aggregation result can then be included as a separate foreground node with a unit weight, and the aggregated inventory vector included as the corresponding column of $B_f$.  This is illustrated in Fig.~\ref{fig:private}c.

The author may relax the disclosure constraints by reporting the locations of some or all non-zero entries in $A_{d,priv}$ and $B_{f,priv}$ but not disclosing their values, or disclosing a range that includes the actual value.  This would enable a reader to understand \emph{that} a certain background process or emission was included in the model (and checked by the critical reviewer) without knowing how much, which may jointly satisfy review objectives and disclosure constraints in some cases.

### 2017-11-04 aggregation/agg-review.tex
Late afternoon read through

>>>>

Moreover, because the disclosure of the PSM includes only references to background data without including the data themselves, it is trivial for item~\ref{itm:bg} to include references to a variety of background databases without adding any additional requirements.  For instance, a study that makes reference to a set of EPDs published by one firm, while using a background LCI database from another firm, can express both dependencies in an analogous way, even if the two sources are published independently.
